Repeatability of Pre-Flashover Fire Patterns on Gypsum Wall Board

### 2015-03-07 Projects/Madrzykowski_Thesis/Report/Madrzykowski_Thesis.tex
Madrzy_Thesis: update directory structure
Madrzy_Thesis: minor formatting updates

\usepackage{graphicx}

@@

\abstract{%
Uncontrolled fires result in loss of life, property and can create an adverse economic impact on a community.  The investigation of fires provides a means to identify the cause of the fire in order to develop a knowledge base that could enable the elimination of that cause and thus reduce the losses from fires.  However, as the result of the review of several arson cases and a review by the U.S. National Academy of Sciences, questions about the lack of the use of science in the practice of fire investigation were raised.  Specifically the National Academy of Sciences indicated that ``... research is needed on the natural variability of burn patterns...''
This study addresses that need in two ways:
\begin{enumerate} 
\item Examining the repeatability of several small fire sources and the fire patterns that were generated by those fires 
\item Examining the capability of using numerical models to simulate the fires and the resulting fire patterns based on input data collected from engineering reference sources, and bench scale and full scale fire experiments conducted in this study.
\end{enumerate}
This paper highlights the many uncertainties involved in what appear to be simple fire experiments.  A variety of uncertainties related to the composition of the fuels, measurement methods, and analysis techniques just to list a few of the areas considered.  The objective of the research was to examine the repeatability of pre-flashover fire patterns generated from exposure to short duration (300 s maximum) well characterized fires.  Three different fuels were used: natural gas, gasoline, and polyurethane foam.  Each fuel had a similar top surface area.  The heat release rate data showed that the variability was greater for more complex fuels.  While the variation in peak heat release rate with the natural gas was similar to the expanded uncertainty of the measurement system, 11~\%, the variation in the peak heat release rates of the gasoline and the polyurethane foam increased due to increased uncertainties in the burning behavior of the fuels.  The patterns generated by the polyurethane foam fire, had more variability than the natural gas and the gasoline fire patterns.  This is likely due to the greater variability of the burning of the solid fuel and the lower heat release rate.  The maximum fire pattern heights generated from the natural gas and gasoline fires were shown to have uncertainties of 18~\% or less based on a Type A statistical analysis with 95~\% confidence limits.  The comparison of the fire pattern heights and the mean flame height demonstrated that the “steady state” natural gas fires exhibited the highest level of agreement and polyurethane foam fueled fire exhibited the least agreement, with the gasoline fueled fires in between.

@@

\acknowledgments{%
I would like to thank Prof.~Fleishmann, etc.

@@

Uncontrolled fires result in loss of life, property and can create an adverse economic impact on a community.  Findings from an archeological site in South Africa, published in 2012, indicate that mankind has productively used fire for approximately 1 million years.[1]  This is 300 thousand years earlier than previously thought.  Despite mankind’s extensive history with fire, according to the World Health Organization, an estimated 265,000 people die each year as a result of burn injuries.[2]  Based on an estimated global population of 7.3 billion people, the death rate due to burns from all sources would be 3.6 per 100,000 populations.  The World Fire Statistics Center collects data from European, North American and a few Asia-Oceania countries.  Within this group of countries the fire deaths per 100,000 population range from 0.02 (Singapore) to 2.03 (Finland). [3]      There are a significant number of factors that determine the variability of fire death rates between nations, socio- economic conditions being a leading factor.[2]

@@

Patterns produced by the fire are in many cases a significant portion of the data collected and are analyzed at the scene to determine the area of origin.  One of the basic methods of documenting the fire scene is to photograph fire patterns.  As noted by Icove, DeHaan, and Haynes  “the ability to document and interpret fire patterns accurately is essential to investigators reconstructing fire scenes…”. [13]  A fire pattern is defined in NFPA 921 as, “the visible or measurable physical change, or identifiable shapes, formed by a fire effect or group of fire effects” [921].  
DeHaan [12] categorizes the fire effects that form patterns as follows:
\begin{enumerate}
\item Surface deposits – no irreversible effect on surface
\item Surface thermal effects – physical change such a discoloration or melting
\item Charring –evidence of surface burning
\item Penetration – charring below the surface
\item Consumption –loss of surface material, charring throughout
\end{enumerate}
Lentini points out that most fire patterns are generated and later observed on two-dimensional surfaces at the place where those surfaces intersect with the three-dimensional fire. Various [14] types of fire patterns, such as; “V-shaped”, “hour-glass”, and “inverted cone”, have come from common observation at actual fire scenes.  As a result, the observations are typically qualitative in nature.   
Previous fire pattern research by the National Institute of Justice (NIJ), the National Institute of Standards and Technology (NIST), and the United States Fire Administration (USFA) has shown that fire patterns provide data useful for the determination of fire origin.  The reports noted the impact of ventilation on the development of the burn patterns. [16, 17] A large number of other factors affect the formation of these patterns: burn time, heat release rate of the fire source, fire exposure, target fuel composition, adjacent fuel(s) and compartmentation, to name a few.    Given the limited number of experiments in the literature and the large number of variables, it has been difficult to fully understand a cause and effect relationship between the fire scenarios and the resulting patterns.  Therefore, fire models are another tool that may be used by a fire investigation team to gain insight into the growth and development of a fire and to test the investigator’s hypothesis on origin and cause.[18]  Many theoretical and empirical correlations and computational fire models have been developed for use in fire protection design, fire hazard analysis, and fire investigation.  A collection of the widely used correlations can be found in the U.S. Nuclear Regulatory Commission’s (NRC) Fire Dynamics Tools (FDTs).  FDTs incorporates references to the source documents and experimental work upon which the correlations are based. In addition, NRC tasked NIST with conducting a verification and validation study of the correlations. Overholt developed the NIST verification and validation study.[19] In a private communication, Overholt noted that not all of the correlations in FDTs were included in the study since a sufficient amount of reliable data in the literature to not available to conduct validation.  Two of the types of correlations, which could not be validated, were flame height and time to ignition.[19]

@@

Of the active field models, FDS version 6 has one of the most rigorous and publically accessible verification and validation programs.[22] In the current FDS validation guide more than 45 experimental data sets are used to quantify the prediction capabilities of the model.  Results are presented in terms of model bias and variation to inform users of whether particular quantities (temperature, heat flux, pressure, etc.) are over or under predicted compared to experimental data.

However for fire models to be demonstrated as reliable and suitable for use in court, studies such as the one conducted here need to be conducted to provide the scientific basis for the use of fire models for re-construction.  The research results may enable fire investigation teams to have a better understanding of the fire scene and enable improved support of their hypothesis on the area of origin based on fire patterns.  This idea was discussed when the Fire Protection Research Foundation convened a Research Advisory Council on Post-fire Analysis in 2002.  Recommendations for research and development in their white paper included, “advance the capabilities of computational fluid dynamics fire modeling, particularly as applied to fire ignition scenarios and fire pattern development and interpretation”  [23]

It has been pointed out in this introduction that fire is still a threat to life and property and how fire investigation plays a role in preventing future fires.  Given that fire pattern interpretation and analysis is a critical step in most fire investigations, it is surprising to find the limited amount or research and analysis tools directly related to this practice.  The fire investigation community recognized a need to improve the science and practice of fire investigation.  One of the manifestations of the efforts to improve the practice was the development of NFPA 921, Guide for Fire and Explosion Investigations.  The first edition of NFPA 921 was issued in 1992.  As provided by the document scope, it “… is designed to assist individuals who are charged with the responsibility of investigating and analyzing fire and explosion incidents and rendering opinions as to the origin, cause, responsibility, or prevention of such incidents, and the damage and injuries which arise from such incidents.”     The document’s purpose includes the goal to be “…a model for the advancement and practice of fire and explosion investigation, fire science, and methodology.” [8] However, NFPA 921 can only incorporate the fire science and research results that have been produced.  Although NFPA 921 is a “guide” as opposed to a “standard” is considered the standard of care for the fire investigation community in countries that recognize and use NFPA standards and guides.

As the result of the review of several arson cases, most notably the cases of Ernest Ray Willis and Cameron Todd Willingham, the practice of fire investigation as a forensic science was called into question.  Beyler conducted a review of both cases for the Texas Forensic Science Commission.  Both fires occurred prior to the release of NFPA 921 in 1992.  The Beyler analysis, issued in August of 2009, concluded that the investigations of both cases “did not comport with either the modern standard of care expressed by NFPA 921, or the standard of care expressed by fire investigation texts and papers in the period 1980 – 1992.”  The fire investigators had a poor understanding of science and their findings of arson could not be sustained.[24]
Another landmark document regarding the state of forensic science in the United States was published in 2009.  The 328 page report, titled Strengthening Forensic Science in the United States: A Path Forward, by the U.S. National Academy of Sciences, reviewed 13 different forensic science disciplines including; biological evidence, friction ridge patterns, tool mark and firearms identification, and analysis of explosives evidence and fire debris, to name a few.  The review of the analysis of explosives evidence and fire debris is covered in three pages of the report.  The summary assessment of analysis of explosives evidence is positive, “the scientific foundations exist to support the analysis of explosions, because such analysis is based primarily on well-established chemistry .”[25]  The complete summary assessment for fire debris analysis is included here: “By contrast, much more research is needed on the natural variability of burn patterns and damage characteristics and how they are affected by the presence of various accelerants.  Despite the paucity of research, some arson investigators continue to make determinations about whether or not a particular fire was set.  However, according to testimony presented to the committee, many of the rules of thumb, that were typically assumed to indicate that an accelerant was used (e.g. “alligatoring” of wood, specific char patterns) have been shown not to be true.  Experiments should be designed to put arson investigations on a more solid scientific footing .” [25]   
This study is in response to the assessment above, specifically regarding the natural variability of fire patterns.

@@

Given that the majority of intentional fires in structures occurred in residential occupancies, this study will focus on fire patterns on painted gypsum wallboard, which is 2.4 m in height.  Given the upper bound of 2.4 m, the experiments will use small fires with mean flame heights of approximately 1.2 m.  The resulting patterns on the gypsum board wall would be representative of a pre-flashover fire pattern.

@@

The fires were characterized in terms of heat release rate, mass loss rate, fire plume temperatures and emitted heat flux. Videos were recorded for each experiment.  This “field” data from the videos was  used to determine the flame heights. Experiments were conducted with three different fuel types: a gas fueled mass flow controlled burner, a liquid fueled burner, and a solid fuel.
Instrumented Calibration Target Wall – A free standing, non-combustible, calcium silicate wall was instrumented to measure the thermal exposure of the wall from the fire in terms of temperature and heat flux.  The fires were initially placed at predetermined, fixed positions away from the wall. The burner was moved toward the wall until the burner was against the wall.

@@

Comparative experiments were conducted to examine the impact of the method of construction and support of the walls and corner sections on the development of the burn patterns.  Three cases were examined: 1) 12.7 mm thick, regular core, painted gypsum board on the front of a wall frame (open back), 2) 12.7 mm thick, regular core, painted gypsum board on the front and back of a wall frame (interior wall), and 3) 12.7 mm thick, regular core, painted gypsum board on the front and back of a wall frame, with fiberglass batting insulation in the voids of the wood frame (exterior wall).  The main consideration was to identify any differences in the height, width, shape, and area of the fire patterns. The results of these experiments determined the construction of the walls and corners used in subsequent phases of this study.

@@

Each of the test compartments consisted of three 3.6 m long by 2.4 m high wood framed walls. A partial ceiling was installed over the wall. The interior of the wall sections and ceilings were covered with painted gypsum wallboard.  These experiments were conducted without instrumentation in the walls.  Based on the outcome of the instrumented calibration target wall results, burner positions were chosen. After each burn, the fire patterns were measured, photographed and analyzed for repeatability. The fire damaged wallboard was subsequently removed and replaced with new sections of painted gypsum wallboard. The number of experiments for each case was dependent on the repeatability and similarity of the pattern development.

@@

Several types of bench scale experiments were conducted to generate data. This data could be used to assist in the analysis of the fire patterns and potentially be used as input for the FDS models.
Critical Heat Flux – Conducted a series of experiments on samples of painted gypsum wallboard to determine the repeatability of a “critical heat flux” that results in a clear line of demarcation, a “burn/no burn” line.  The flooring radiant panel (ASTM E648) was used to examine this value.

@@

Cone calorimeter (ASTM 1354) experiments were conducted to estimate the ignition temperature, the time to ignition and the heat release rate from samples of painted gypsum wallboard under thermal flux conditions ranging from 25 kW/m2 to 75 kW/m2.

@@

A series of experiments were conducted to characterize a limited set of  residential scale, pre-flashover or incipient fires.  A fixed fire source footprint, 30.5 cm square, was chosen for this study and a range of common fuels was selected.  The fires were positioned under an oxygen consumption calorimetry hood to determine the heat release rate.  Heat release rate, flame/plume gas temperatures, and total heat flux from the flames were measured and flame movement and height were recorded with photographs and videos.  Given the focus on residential fires, small source fires were selected so that the burn pattern would be limited to less than 2.4 m in height.  Replicate experiments were conducted with each of the fuels, in order to examine the repeatability of the fires and to quantify the variability in the flames.

@@

The natural gas burner was 30.5 cm on a side and the top surface of the burner was 7.5 cm above the floor. The shell of the burner was made from steel and filled with “pea sized” gravel with an average diameter of 6 mm.  A steady state heat release rate of approximately 80 kW was used for the experiments.

Gasoline was the fuel of choice for the liquid fueled pan fire.  Gasoline is not an ideal fuel to use in experiments because it is a mixture of components and additives that are specific to manufacturers. In addition the mixtures change from season to season, and vary based on requirements or restrictions of the locality.  However, fire incident data shows that gasoline was the first item ignited in the majority of intentional fires where flammable or combustible liquids were used 12.[30] In addition, data from forensic laboratories, collected by Babrauskas and others, indicates that gasoline is the most prevalent accelerant found as part of the analysis of fire debris samples 13, 14.[31, 32] 
The gasoline pan fire used 500 mL of unleaded, 87 octane gasoline.  The pan was constructed from 6 mm thick steel.  The inner dimension of the pan was 30.5 cm on a side and 1.9 cm high.  The pan was elevated in order to bring the pan lip height up to 7.5 cm above the floor to be the same height as the natural gas burner.  The 500 mL of gasoline had a fuel depth of 5.5 mm and provided a burn time of approximately 300 s.

@@

The flexible polyurethane foam had a density of approximately 23 kg/m3.  Each block of foam was 30.5 cm on a side and was 7.5 cm thick. The bottom and sides of the foam block were wrapped in aluminum foil to prevent the fuel from flowing or spreading as it burned to ensure that the fuel area remained with the same fixed area as the gasoline and natural gas.

@@

The source fire characterization experiments were conducted in the NIST Large Fire Research Laboratory’s 3 m by 3 m hood oxygen consumption calorimeter. This apparatus follows the methodology of ASTM E 2067-08, Standard Practice for Full-Scale Oxygen Consumption Calorimetry Fire Tests.[37]  The fires were centered under the hood of the calorimeter and positioned in an open steel frame which provided support for the thermocouples and heat flux sensors. A load cell was positioned as a support for the fuels.  All of the data was recorded at intervals of 1 s on a computer based data acquisition system.

@@

The NIST Large Fire Research Laboratory’s 3 m by 3 m oxygen depletion calorimeter has an estimated peak heat release rate capacity of approximately 700 kW.  Details on the operation and uncertainty in measurements associated with NIST’s 6 m by 6 m oxygen consumption calorimeter were documented by Bryant et.al. [38] Many of the same instruments are shared between the two calorimeters and calibration burns were conducted during each experimental series.  As a result, the expanded uncertainty of this device has been estimated as 11~\% on the measured heat release rate.

@@

Gas temperatures were measured with exposed-bead, Chromel-Alumel (type K) thermocouples, with a 1.0 mm nominal diameter.  Starting from the exposed bead, the thermocouple wire was sheathed in a 3.2 mm diameter inconel shield, 0.76 m in length.  The standard uncertainty in temperature of the thermocouple wire itself is 2.2~$^\circ$C at 277~$^\circ$C and increases to 9.5~$^\circ$C at 871~$^\circ$C as determined by the wire manufacturer [39]  The variation of the temperature in the environment surrounding the thermocouple is known to be much greater than that of the wire uncertainty.[40, 41]  Small diameter thermocouples were used to limit the impact of radiative heating and cooling.  The estimated combined expanded uncertainty for temperature in these experiments is 15~\%.

@@

Water cooled, Schmidt-Boelter total heat flux gauges were used to measure the heat flux. The sensing surface of the heat flux gauges were located 0.95 m horizontally from the center of the burner/fuel and 0.50 m above the top of the burner/pan lip/fuel surface.  The manufacturer reports a 3~\% calibration expanded uncertainty for these devices 20.[42] Results from an international study on total heat flux gauge calibration and response demonstrated that the uncertainty of a Schmidt-Boelter gauge is typically 8~\%.[43]  The gauges were mounted to uprights of the steel instrumentation support frame on opposite sides of the fire.

@@

The fires were centered under the hood of the calorimeter on a non-combustible platform measuring 0.91 m x 1.2 m x 12 mm thick supported by the load cell.  The load cell had a resolution of 1.0 g.[44] The expanded uncertainty is estimated to be 5~\%.  The load cell was checked between experiments with calibrated weights during the zero and span checks of the calorimeter.

@@

A 35 mm single lens reflex camera was used to record digital still photographs of the fires.  The camera had an 8 megapixel resolution.  A mini DV camcorder with 0.7 megapixel resolution was used to capture the video.  A firefighting type thermal imager with 0.08 megapixel resolution was used to capture infrared images in the 8 to 14 micron range.  Each camera was positioned on a tripod with a distance of 4.5 m between the centerline of burner to camera lens 4.5 m.  The height of each camera was adjusted so that the center of camera lens was positioned 0.5 m above top of the burner.

A schematic drawing of the instrumentation arrangement is shown in Fig.~\ref{Drawing_Fire_Flamer}.  The thermocouples were mounted on a steel frame.  The frame enabled the thermocouples to be installed at 10 cm intervals from 10 cm to 120 cm above the burner/pan lip/fuel surface. At each 10 cm interval, there were five thermocouples: one centered above the burner/fuel surface and one centered above each edge of the burner/fuel surface, for a total of 70 thermocouples.  A plumb bob and steel tape measure, with a resolution of $\pm$ 0.5 mm, were used to position and align the thermocouples.   The expanded uncertainty for the location of each thermocouple tip is estimated to be $\pm$ 2.5~\%.  The same applies to the location of the heat flux sensor surfaces.

>>>>

\usepackage{graphicx}
\frenchspacing

@@

\abstract{
Uncontrolled fires result in loss of life, property and can create an adverse economic impact on a community.  The investigation of fires provides a means to identify the cause of the fire in order to develop a knowledge base that could enable the elimination of that cause and thus reduce the losses from fires.  However, as the result of the review of several arson cases and a review by the U.S. National Academy of Sciences, questions about the lack of the use of science in the practice of fire investigation were raised.  Specifically the National Academy of Sciences indicated that ``... research is needed on the natural variability of burn patterns...''
This study addresses that need in two ways:
\begin{enumerate} 
\item Examining the repeatability of several small fire sources and the fire patterns that were generated by those fires 
\item Examining the capability of using numerical models to simulate the fires and the resulting fire patterns based on input data collected from engineering reference sources, and bench scale and full scale fire experiments conducted in this study.
\end{enumerate}
This paper highlights the many uncertainties involved in what appear to be simple fire experiments.  A variety of uncertainties related to the composition of the fuels, measurement methods, and analysis techniques just to list a few of the areas considered.  The objective of the research was to examine the repeatability of pre-flashover fire patterns generated from exposure to short duration (300 s maximum) well characterized fires.  Three different fuels were used: natural gas, gasoline, and polyurethane foam.  Each fuel had a similar top surface area.  The heat release rate data showed that the variability was greater for more complex fuels.  While the variation in peak heat release rate with the natural gas was similar to the expanded uncertainty of the measurement system, 11~\%, the variation in the peak heat release rates of the gasoline and the polyurethane foam increased due to increased uncertainties in the burning behavior of the fuels.  The patterns generated by the polyurethane foam fire, had more variability than the natural gas and the gasoline fire patterns.  This is likely due to the greater variability of the burning of the solid fuel and the lower heat release rate.  The maximum fire pattern heights generated from the natural gas and gasoline fires were shown to have uncertainties of 18~\% or less based on a Type A statistical analysis with 95~\% confidence limits.  The comparison of the fire pattern heights and the mean flame height demonstrated that the ``'steady state'' natural gas fires exhibited the highest level of agreement and polyurethane foam fueled fire exhibited the least agreement, with the gasoline fueled fires in between.

@@

\acknowledgments{
I would like to thank Prof.~Fleishmann, etc.

@@

Uncontrolled fires result in loss of life, property and can create an adverse economic impact on a community.  Findings from an archaeological site in South Africa, published in 2012, indicate that mankind has productively used fire for approximately 1 million years.[1]  This is 300 thousand years earlier than previously thought.  Despite mankind’s extensive history with fire, according to the World Health Organization, an estimated 265,000 people die each year as a result of burn injuries.[2]  Based on an estimated global population of 7.3 billion people, the death rate due to burns from all sources would be 3.6 per 100,000 populations.  The World Fire Statistics Center collects data from European, North American and a few Asia-Oceania countries.  Within this group of countries the fire deaths per 100,000 population range from 0.02 (Singapore) to 2.03 (Finland). [3] There are a significant number of factors that determine the variability of fire death rates between nations, socio- economic conditions being a leading factor.[2]

@@

Patterns produced by the fire are in many cases a significant portion of the data collected and are analyzed at the scene to determine the area of origin.  One of the basic methods of documenting the fire scene is to photograph fire patterns.  As noted by Icove, DeHaan, and Haynes  ``'the ability to document and interpret fire patterns accurately is essential to investigators reconstructing fire scenes…''. [13]  A fire pattern is defined in NFPA 921 as, “the visible or measurable physical change, or identifiable shapes, formed by a fire effect or group of fire effects” [921].  
DeHaan [12] categorizes the fire effects that form patterns as follows:
\begin{enumerate}
\item Surface deposits – no irreversible effect on surface
\item Surface thermal effects – physical change such a discoloration or melting
\item Charring –evidence of surface burning
\item Penetration – charring below the surface
\item Consumption –loss of surface material, charring throughout
\end{enumerate}
Lentini points out that most fire patterns are generated and later observed on two-dimensional surfaces at the place where those surfaces intersect with the three-dimensional fire. Various [14] types of fire patterns, such as; ``V-shaped'', ``hour-glass'', and ``inverted cone'', have come from common observation at actual fire scenes.  As a result, the observations are typically qualitative in nature.

Previous fire pattern research by the National Institute of Justice (NIJ), the National Institute of Standards and Technology (NIST), and the United States Fire Administration (USFA) has shown that fire patterns provide data useful for the determination of fire origin.  The reports noted the impact of ventilation on the development of the burn patterns. [16, 17] A large number of other factors affect the formation of these patterns: burn time, heat release rate of the fire source, fire exposure, target fuel composition, adjacent fuel(s) and compartmentation, to name a few.    Given the limited number of experiments in the literature and the large number of variables, it has been difficult to fully understand a cause and effect relationship between the fire scenarios and the resulting patterns.  Therefore, fire models are another tool that may be used by a fire investigation team to gain insight into the growth and development of a fire and to test the investigator’s hypothesis on origin and cause.[18]  Many theoretical and empirical correlations and computational fire models have been developed for use in fire protection design, fire hazard analysis, and fire investigation.  A collection of the widely used correlations can be found in the U.S. Nuclear Regulatory Commission’s (NRC) Fire Dynamics Tools (FDTs).  FDTs incorporates references to the source documents and experimental work upon which the correlations are based. In addition, NRC tasked NIST with conducting a verification and validation study of the correlations. Overholt developed the NIST verification and validation study.[19] In a private communication, Overholt noted that not all of the correlations in FDTs were included in the study since a sufficient amount of reliable data in the literature to not available to conduct validation.  Two of the types of correlations, which could not be validated, were flame height and time to ignition.[19]

@@

Of the active field models, FDS version 6 has one of the most rigorous and publicly accessible verification and validation programs.[22] In the current FDS validation guide more than 45 experimental data sets are used to quantify the prediction capabilities of the model.  Results are presented in terms of model bias and variation to inform users of whether particular quantities (temperature, heat flux, pressure, etc.) are over or under predicted compared to experimental data.

However for fire models to be demonstrated as reliable and suitable for use in court, studies such as the one conducted here need to be conducted to provide the scientific basis for the use of fire models for re-construction.  The research results may enable fire investigation teams to have a better understanding of the fire scene and enable improved support of their hypothesis on the area of origin based on fire patterns.  This idea was discussed when the Fire Protection Research Foundation convened a Research Advisory Council on Post-fire Analysis in 2002.  Recommendations for research and development in their white paper included, ``advance the capabilities of computational fluid dynamics fire modeling, particularly as applied to fire ignition scenarios and fire pattern development and interpretation.''  [23]

It has been pointed out in this introduction that fire is still a threat to life and property and how fire investigation plays a role in preventing future fires.  Given that fire pattern interpretation and analysis is a critical step in most fire investigations, it is surprising to find the limited amount or research and analysis tools directly related to this practice.  The fire investigation community recognized a need to improve the science and practice of fire investigation.  One of the manifestations of the efforts to improve the practice was the development of NFPA 921, Guide for Fire and Explosion Investigations.  The first edition of NFPA 921 was issued in 1992.  As provided by the document scope, it `` is designed to assist individuals who are charged with the responsibility of investigating and analyzing fire and explosion incidents and rendering opinions as to the origin, cause, responsibility, or prevention of such incidents, and the damage and injuries which arise from such incidents.'' The document’s purpose includes the goal to be ``…a model for the advancement and practice of fire and explosion investigation, fire science, and methodology.'' [8] However, NFPA 921 can only incorporate the fire science and research results that have been produced.  Although NFPA 921 is a ``guide'' as opposed to a ``standard'' is considered the standard of care for the fire investigation community in countries that recognize and use NFPA standards and guides.

As the result of the review of several arson cases, most notably the cases of Ernest Ray Willis and Cameron Todd Willingham, the practice of fire investigation as a forensic science was called into question.  Beyler conducted a review of both cases for the Texas Forensic Science Commission.  Both fires occurred prior to the release of NFPA 921 in 1992.  The Beyler analysis, issued in August of 2009, concluded that the investigations of both cases ``did not comport with either the modern standard of care expressed by NFPA 921, or the standard of care expressed by fire investigation texts and papers in the period 1980 – 1992.'' The fire investigators had a poor understanding of science and their findings of arson could not be sustained.[24]

Another landmark document regarding the state of forensic science in the United States was published in 2009.  The 328 page report, titled Strengthening Forensic Science in the United States: A Path Forward, by the U.S. National Academy of Sciences, reviewed 13 different forensic science disciplines including; biological evidence, friction ridge patterns, tool mark and firearms identification, and analysis of explosives evidence and fire debris, to name a few.  The review of the analysis of explosives evidence and fire debris is covered in three pages of the report.  The summary assessment of analysis of explosives evidence is positive, ``the scientific foundations exist to support the analysis of explosions, because such analysis is based primarily on well-established chemistry .''[25]  The complete summary assessment for fire debris analysis is included here: ``By contrast, much more research is needed on the natural variability of burn patterns and damage characteristics and how they are affected by the presence of various accelerants.  Despite the paucity of research, some arson investigators continue to make determinations about whether or not a particular fire was set.  However, according to testimony presented to the committee, many of the rules of thumb, that were typically assumed to indicate that an accelerant was used (e.g. ``alligatoring'' of wood, specific char patterns) have been shown not to be true.  Experiments should be designed to put arson investigations on a more solid scientific footing .'' [25]   
This study is in response to the assessment above, specifically regarding the natural variability of fire patterns.

@@

Given that the majority of intentional fires in structures occurred in residential occupancies, this study will focus on fire patterns on painted gypsum wallboard, which is 2.4~m in height.  Given the upper bound of 2.4~m, the experiments will use small fires with mean flame heights of approximately 1.2~m.  The resulting patterns on the gypsum board wall would be representative of a pre-flashover fire pattern.

@@

The fires were characterized in terms of heat release rate, mass loss rate, fire plume temperatures and emitted heat flux. Videos were recorded for each experiment.  This ``field'' data from the videos was  used to determine the flame heights. Experiments were conducted with three different fuel types: a gas fueled mass flow controlled burner, a liquid fueled burner, and a solid fuel.
Instrumented Calibration Target Wall – A free standing, non-combustible, calcium silicate wall was instrumented to measure the thermal exposure of the wall from the fire in terms of temperature and heat flux.  The fires were initially placed at predetermined, fixed positions away from the wall. The burner was moved toward the wall until the burner was against the wall.

@@

Comparative experiments were conducted to examine the impact of the method of construction and support of the walls and corner sections on the development of the burn patterns.  Three cases were examined: 1) 12.7~mm thick, regular core, painted gypsum board on the front of a wall frame (open back), 2) 12.7~mm thick, regular core, painted gypsum board on the front and back of a wall frame (interior wall), and 3) 12.7~mm thick, regular core, painted gypsum board on the front and back of a wall frame, with fiberglass batting insulation in the voids of the wood frame (exterior wall).  The main consideration was to identify any differences in the height, width, shape, and area of the fire patterns. The results of these experiments determined the construction of the walls and corners used in subsequent phases of this study.

@@

Each of the test compartments consisted of three 3.6~m long by 2.4~m high wood framed walls. A partial ceiling was installed over the wall. The interior of the wall sections and ceilings were covered with painted gypsum wallboard.  These experiments were conducted without instrumentation in the walls.  Based on the outcome of the instrumented calibration target wall results, burner positions were chosen. After each burn, the fire patterns were measured, photographed and analyzed for repeatability. The fire damaged wallboard was subsequently removed and replaced with new sections of painted gypsum wallboard. The number of experiments for each case was dependent on the repeatability and similarity of the pattern development.

@@

Several types of bench scale experiments were conducted to generate data. This data could be used to assist in the analysis of the fire patterns and potentially be used as input for the FDS models.
Critical Heat Flux – Conducted a series of experiments on samples of painted gypsum wallboard to determine the repeatability of a “critical heat flux” that results in a clear line of demarcation, a ``burn/no burn'' line.  The flooring radiant panel (ASTM E648) was used to examine this value.

@@

Cone calorimeter (ASTM 1354) experiments were conducted to estimate the ignition temperature, the time to ignition and the heat release rate from samples of painted gypsum wallboard under thermal flux conditions ranging from 25~kW/m$^2$ to 75~kW/m$^2$.

@@

A series of experiments were conducted to characterize a limited set of  residential scale, pre-flashover or incipient fires.  A fixed fire source footprint, 30.5~cm square, was chosen for this study and a range of common fuels was selected.  The fires were positioned under an oxygen consumption calorimetry hood to determine the heat release rate.  Heat release rate, flame/plume gas temperatures, and total heat flux from the flames were measured and flame movement and height were recorded with photographs and videos.  Given the focus on residential fires, small source fires were selected so that the burn pattern would be limited to less than 2.4~m in height.  Replicate experiments were conducted with each of the fuels, in order to examine the repeatability of the fires and to quantify the variability in the flames.

@@

The natural gas burner was 30.5 cm on a side and the top surface of the burner was 7.5~cm above the floor. The shell of the burner was made from steel and filled with “pea sized” gravel with an average diameter of 6~mm.  A steady state heat release rate of approximately 80 kW was used for the experiments.

Gasoline was the fuel of choice for the liquid fueled pan fire.  Gasoline is not an ideal fuel to use in experiments because it is a mixture of components and additives that are specific to manufacturers. In addition the mixtures change from season to season, and vary based on requirements or restrictions of the locality.  However, fire incident data shows that gasoline was the first item ignited in the majority of intentional fires where flammable or combustible liquids were used 12.[30] In addition, data from forensic laboratories, collected by Babrauskas and others, indicates that gasoline is the most prevalent accelerant found as part of the analysis of fire debris samples 13, 14.[31, 32] 
The gasoline pan fire used 500 mL of unleaded, 87 octane gasoline.  The pan was constructed from 6~mm thick steel.  The inner dimension of the pan was 30.5 cm on a side and 1.9~cm high.  The pan was elevated in order to bring the pan lip height up to 7.5~cm above the floor to be the same height as the natural gas burner.  The 500~mL of gasoline had a fuel depth of 5.5~mm and provided a burn time of approximately 300~s.

@@

The flexible polyurethane foam had a density of approximately 23~kg/m$^3$.  Each block of foam was 30.5~cm on a side and was 7.5~cm thick. The bottom and sides of the foam block were wrapped in aluminum foil to prevent the fuel from flowing or spreading as it burned to ensure that the fuel area remained with the same fixed area as the gasoline and natural gas.

@@

The source fire characterization experiments were conducted in the NIST Large Fire Research Laboratory’s 3~m by 3~m hood oxygen consumption calorimeter. This apparatus follows the methodology of ASTM E 2067-08, Standard Practice for Full-Scale Oxygen Consumption Calorimetry Fire Tests.[37]  The fires were centered under the hood of the calorimeter and positioned in an open steel frame which provided support for the thermocouples and heat flux sensors. A load cell was positioned as a support for the fuels.  All of the data was recorded at intervals of 1~s on a computer based data acquisition system.

@@

The NIST Large Fire Research Laboratory’s 3~m by 3~m oxygen depletion calorimeter has an estimated peak heat release rate capacity of approximately 700~kW.  Details on the operation and uncertainty in measurements associated with NIST’s 6~m by 6~m oxygen consumption calorimeter were documented by Bryant et al. [38] Many of the same instruments are shared between the two calorimeters and calibration burns were conducted during each experimental series.  As a result, the expanded uncertainty of this device has been estimated as 11~\% on the measured heat release rate.

@@

Gas temperatures were measured with exposed-bead, Chromel-Alumel (type K) thermocouples, with a 1.0~mm nominal diameter.  Starting from the exposed bead, the thermocouple wire was sheathed in a 3.2 mm diameter inconel shield, 0.76~m in length.  The standard uncertainty in temperature of the thermocouple wire itself is 2.2~$^\circ$C at 277~$^\circ$C and increases to 9.5~$^\circ$C at 871~$^\circ$C as determined by the wire manufacturer [39]  The variation of the temperature in the environment surrounding the thermocouple is known to be much greater than that of the wire uncertainty.[40, 41]  Small diameter thermocouples were used to limit the impact of radiative heating and cooling.  The estimated combined expanded uncertainty for temperature in these experiments is 15~\%.

@@

Water cooled, Schmidt-Boelter total heat flux gauges were used to measure the heat flux. The sensing surface of the heat flux gauges were located 0.95~m horizontally from the center of the burner/fuel and 0.50~m above the top of the burner/pan lip/fuel surface.  The manufacturer reports a 3~\% calibration expanded uncertainty for these devices 20.[42] Results from an international study on total heat flux gauge calibration and response demonstrated that the uncertainty of a Schmidt-Boelter gauge is typically 8~\%.[43]  The gauges were mounted to uprights of the steel instrumentation support frame on opposite sides of the fire.

@@

The fires were centered under the hood of the calorimeter on a non-combustible platform measuring 0.91~m x 1.2~m x 12~mm thick supported by the load cell.  The load cell had a resolution of 1.0~g.[44] The expanded uncertainty is estimated to be 5~\%.  The load cell was checked between experiments with calibrated weights during the zero and span checks of the calorimeter.

@@

A 35~mm single lens reflex camera was used to record digital still photographs of the fires.  The camera had an 8 megapixel resolution.  A mini DV camcorder with 0.7 megapixel resolution was used to capture the video.  A firefighting type thermal imager with 0.08 megapixel resolution was used to capture infrared images in the 8 to 14 micron range.  Each camera was positioned on a tripod with a distance of 4.5 m between the centerline of burner to camera lens 4.5~m.  The height of each camera was adjusted so that the center of camera lens was positioned 0.5 m above top of the burner.

A schematic drawing of the instrumentation arrangement is shown in Fig.~\ref{Drawing_Fire_Flamer}.  The thermocouples were mounted on a steel frame.  The frame enabled the thermocouples to be installed at 10~cm intervals from 10~cm to 120~cm above the burner/pan lip/fuel surface. At each 10~cm interval, there were five thermocouples: one centered above the burner/fuel surface and one centered above each edge of the burner/fuel surface, for a total of 70 thermocouples.  A plumb bob and steel tape measure, with a resolution of $\pm$ 0.5~mm, were used to position and align the thermocouples.   The expanded uncertainty for the location of each thermocouple tip is estimated to be $\pm$ 2.5~\%.  The same applies to the location of the heat flux sensor surfaces.
